\part{DSL Implementation}
 
\vspace{10mm}
\noindent This part of the book has been written together Lennart Kats and Guido
Wachsmuth (who have contributed the material on Spoofax) and Christian Dietrich,
who helped with the language modularization in Xtext.
\vspace{10mm}
 

In this part we describe language implementation with three language workbench
tools which are all Open Source and representative for the current state of the
art of language workbenches: Spoofax, Xtext and MPS. For more example language
implementations using more language workbenches, take a look at the Language
Workbench Competition website at \ic{http://languageworkbenches.net}.

This chapter does not cover a lot of design decisions or motivation for having
things like constraints, type systems, transformations or generators.
Conceptually these topics are introduced in Part II of the book on DSL design.
This part really just looks at the "how", not the "what" or "why".

Each chapter contains examples from all three tools. The ordering of the
tools is different from chapter to chapter based on the characteristics of each
tool: if the example for tool A illustrates a point that is also relevant for
tool B, then A is discussed before B.

The too examples are not intended to serve as a full tutorial for any
of these tools, but as an illustration of the concepts and ideas involved with
language implementation in general. However, they should give you a solid
understanding of the capabilities of each of the tools, and the class of tools
they stand for.





